# Curve fitting, with examples from Red Clover Valley carbon flux and gullies

This section will explore various curve fitting methods, starting with some carbon flux tower data from Red Clover Valley. There's not a lot of interpretation, as this is basically a sandbox for now...


## Red Clover Valley flux data

1. Initial data read and time series charting

Some ideas for processing are from Mason Kealy.  Data are from Andrew Oliphant.

```{r include=FALSE}
library(readxl)
library(tidyverse)
library(lubridate)
```

### Reading in the Excel File

First create a function to convert the first two rows into individual variables. We won't include the units.

```{r}
library(igisci); library(tidyverse); library(readxl)
```
```{r}
read_excel2 <- function(xl){
  vnames <- read_excel(xl, n_max=0) |> names()
  vunits <- read_excel(xl, n_max=0, skip=1) |> names()
  read_excel(xl, skip=2, col_names=vnames)
}
```

In reading the data, we'll remove the zero GPP_sum values. 
```{r}
Carbon_Flux <- read_excel2(ex("meadows/RCV_carbon_flux_data.xlsx")) |>
  filter(GPP_sum>0)
```
  

```{r}
Mountflux <- Carbon_Flux %>% 
mutate_all(as.numeric) %>% 
mutate(DOY = as.Date(Date,format = "%m-%d")) %>% 
select(Date:PAR_ave) %>% 
mutate(NEE = round(NEE_sum, digits = 3), 
       GPP = round(GPP_sum, digits = 3),
       RE = round(RE_sum, digits = 3)) %>%
#select(-NEE_sum, -GPP_sum, -RE_sum) %>% 
  pivot_longer(cols = c("NEE", "RE", "GPP"), names_to = "CO2_FLUX_COMPONENTS", values_to ="value") %>%
  filter(GPP_sum != 0, !is.na(PAR_ave))

print(Mountflux)
```

```{r}

```


### Time series plot 
  Then I used ggplot to create a visualization of how NEE, GPP, an RE changed over a yearly time scale.
```{r}

Mountflux %>% 
 ggplot(aes(x = Date, y= value, color = CO2_FLUX_COMPONENTS))+
  geom_line()+
  geom_hline(yintercept = 0, linetype = "longdash", color = "black")+
  ggtitle(" Red Clover Valley: Interannual Components of Carbon Flux Measurements") +
  xlab("Day of Year") +
  ylab("g C/m^2")+
  labs(caption =
"Figure 1: Components of Carbon Flux measurements collected in 30 minute intervals in 2021
and added into ‘sum daily value’ , from flux tower in Red Clover Valley,California. Components included NEE (Net Ecosystem Exchange), RE (Respiration Efficiency), GPP (Gross Primary Productivity) all measured in g C m^2 and are plotted against an annual time scale with days as the unit of measure.")
  
```
 
### Pairs plot
```{r message=F,warning=F}
Carbon_Flux |>
  select(GPP_sum:vpd_max) |>
  GGally::ggpairs()
  # pairs()
  
```
 

### Non-linear model of PAR_ave ~ GPP_sum, using Michaelis-Menten model

*Not the right goal for a model between these variables, since GPP should derive from PAR, not the other way around.  But it does illustrate asymptotic data.*


First the scatter plot:

```{r}
plot(Carbon_Flux$GPP_sum, Carbon_Flux$PAR_ave,xlab="GPP_sum", ylab="PAR_ave")
```


### Try fitting a non-linear model (SSmicmen from base R)

Based on https://www.datacamp.com/tutorial/introduction-to-non-linear-model-and-insights-using-r

Using start values for Vm and K based on interpreting the graph. We'll use nonlinear least squares (`nls`) and starting values of the two coefficients.

```{r}
micmenMod <- nls(PAR_ave~(Vm)*GPP_sum/(K+GPP_sum), data=Mountflux, start=list(Vm=750,K=1))
summary(micmenMod)
```
**Calculate residual error**

```{r}
sse <- micmenMod$m$deviance()
null <- lm(PAR_ave~1, data=Mountflux)
sst <- data.frame(summary.aov(null)[[1]])$Sum.Sq
percent_variation_explained <- 100*(sst-sse)/sst
percent_variation_explained
```
```{r}
plot(Mountflux$GPP_sum, Mountflux$PAR_ave,xlab="GPP_sum", ylab="PAR_ave")
 lines(Mountflux$GPP_sum, 
       predict(micmenMod,Mountflux),col=2)
```

### polynomial models

Ok, enough of that deceptive model of PAR predicted by GPP -- not the way it works -- and instead we'll look at models that predict GPP from PAR, not asymptotic 

```{r}
plot(Mountflux$PAR_ave,Mountflux$GPP_sum, xlab="PAR_ave", ylab="GPP_sum")

```

#### poly 1

```{r}
ord = 1
summary(lm(GPP_sum ~ poly(PAR_ave, ord, raw=T), data=Mountflux))
ggplot(Mountflux, aes(PAR_ave, GPP_sum)) + geom_point() +
  stat_smooth(method=lm, formula=y~poly(x,ord,raw=T))

```



#### poly 2

```{r}
ord = 2
summary(lm(GPP_sum ~ poly(PAR_ave, ord, raw=T), data=Mountflux))
ggplot(Mountflux, aes(PAR_ave, GPP_sum)) + geom_point() +
  stat_smooth(method=lm, formula=y~poly(x,ord,raw=T))

```


#### poly 3


```{r}
ord = 3
summary(lm(GPP_sum ~ poly(PAR_ave, ord, raw=T), data=Mountflux))
ggplot(Mountflux, aes(PAR_ave, GPP_sum)) + geom_point() +
  stat_smooth(method=lm, formula=y~poly(x,ord,raw=T))
```

#### poly 4
```{r}
ord = 4
summary(lm(GPP_sum ~ poly(PAR_ave, ord, raw=T), data=Mountflux))
ggplot(Mountflux, aes(PAR_ave, GPP_sum)) + geom_point() +
  stat_smooth(method=lm, formula=y~poly(x,ord,raw=T))
```

#### poly 5
```{r}
ord = 5
summary(lm(GPP_sum ~ poly(PAR_ave, ord, raw=T), data=Mountflux))
ggplot(Mountflux, aes(PAR_ave, GPP_sum)) + geom_point() +
  stat_smooth(method=lm, formula=y~poly(x,ord,raw=T))
```

### power law relationship
```{r}
powerModel <- lm(log(GPP_sum) ~ log(PAR_ave), data=Mountflux)
summary(powerModel)
ggplot(Mountflux, aes(log(PAR_ave), log(GPP_sum))) + geom_point() +
  stat_smooth(method=lm)
```
Now plot without the log transformed axes, we'll need to create the curve with the model, using the coefficients.

```{r}
coeff = exp(powerModel$coefficients[1])
power = powerModel$coefficients[2]
fit_func <- function(x) coeff*x^power

fluxPlot <- ggplot(Mountflux) +
  geom_point(aes(x=PAR_ave,y=GPP_sum))
fluxPlot <- fluxPlot +
  stat_function(fun=fit_func, color="orange",lwd=2, alpha=0.5)
fluxPlot
```

```{r}
fit_eq = paste("y = ", round(coeff,6), " x^", round(power,3))
fluxPlot <- fluxPlot +
  annotate("text",x=200,y=40,label=fit_eq,size=4)
fluxPlot
```

## Another curve fit -- of gullies that may follow a rate law of growth, or other asymptotic
from Graf WL (1977). The rate law in fluvial geomorphology. Am. J. Sci 277: 178-191.

The rate-law model that Graf used employs a half life, appropriate when depleting a quantity as will happen with an erosion phase following a disequilibrium (like a geomorphic drop in base level).

```{r}
library(igisci); library(tidyverse)
gullyGrowth <- read_csv(ex("gully/gullyGrowth.csv"))
```

```{r}
ggplot(data=gullyGrowth, aes(x=distance, y=age_yrs, col=type)) +
  geom_point()
```

```{r}
gully <- gullyGrowth |> filter(type=="gully")
gullyMod <- nls(age_yrs~(Vm)*distance/(K+distance), data=gully, start=list(Vm=1,K=1))
summary(gullyMod)
```
**Calculate residual error**

```{r}
sse <- gullyMod$m$deviance()
null <- lm(age_yrs~1, data=gully)
sst <- data.frame(summary.aov(null)[[1]])$Sum.Sq
percent_variation_explained <- 100*(sst-sse)/sst
percent_variation_explained
```

```{r}
proto <- gullyGrowth |> filter(type=="proto-gully")
protoMod <- nls(age_yrs~(Vm)*distance/(K+distance), data=proto, start=list(Vm=1,K=1))
summary(protoMod)
```

```{r}
sse <- protoMod$m$deviance()
null <- lm(age_yrs~1, data=proto)
sst <- data.frame(summary.aov(null)[[1]])$Sum.Sq
percent_variation_explained <- 100*(sst-sse)/sst
percent_variation_explained
```

```{r}
ggplot(data=gullyGrowth, aes(x=distance)) +
  geom_point(aes(y=age_yrs, col=type)) +
  geom_line(data=gully, aes(x=distance, y=predict(gullyMod,gully))) +
  geom_line(data=proto, aes(x=distance, y=predict(protoMod,proto)))
```
