# Figures and Tables

This document demonstrates methods for creating figures and tables suitable for producing nice-looking reports. There are many other resources on this, with probably the most comprehensive being Xie, Allaire, Grolemund (2023). *R Markdown: The Definitive Guide*. https://bookdown.org/yihui/rmarkdown/.  See section 2.6 "R code chunks and inline R code". This document provides a short introduction to these, but the Definitive Guide goes further. Another source, Xie (2023), *bookdown: Authoring Books and Technical Documents with R Markdown* https://bookdown.org/yihui/bookdown/, gets into building chapters, indices, and publishing on bookdown.org.

Note that the methods you use will differ somewhat depending on whether you're outputing to HTML or pdf, and it was a challenge for me figuring out how to do both for the same figure or table.

## Figures

You already know how to create figures in R Markdown, simply by creating graphics in code chunks. To make them more useful as a report, you can add a figure heading and cite the figure in the text. To create the figure heading, assign it in the code chunk header to `fig.cap`. To cite it in text use an inline reference like this: (Figure \@ref(fig:mapsGGlabels)). 

```{r CAmap, warning=F, message=F, fig.cap="California Counties"}
library(igisci); library(ggplot2)
ggplot(CA_counties) + geom_sf() +
  geom_sf_text(aes(label = NAME), size = 1.5)
```

## Tables

There are lots of ways of creating tables in R, some of which are discussed in the book. A recent package from the RStudio/Posit folks is `gt`, worth exploring at https://cloud.r-project.org/web/packages/gt/gt.pdf or https://gt.rstudio.com.

### `DT::datatable`
Here's a table created with the DT package, demonstrating a scroll option useful for interactive documents (HTML).

```{r penguinsDT, out.width="100%", echo=F, tab.cap="DT package datatable output (screen capture)"}
library(palmerpenguins)
DT::datatable(penguins, options=list(scrollX=T))
```

### `knitr::kable` tables

The `knitr` package that RStudio uses to knit documents from R markdown also has a table format function `kable` which has some good options for pdf.

```{r knitrKable}
knitr::kable(head(penguins), caption="a caption")
```

### `gt` package
Here's a table created with the gt package. See the documentation at https://gt.rstudio.com.

```{r include=F}
library(gt); library(tidyverse)
```


```{r}
gt(head(penguins)) |>
  tab_header(title="Palmer Penguins", 
             subtitle="examples of data (first 6 shown)") |>
  tab_footnote("Source: https://allisonhorst.github.io/palmerpenguins/articles/intro.html")
  
```

## Other tricks

It's often useful to include options for the code chunks at the top of the document. To not show this, you can use `echo=F` in the code chunk header.

```{r}
knitr::opts_chunk$set(include=T,echo=T,fig.show=T,warning=F,message=F,fig.align='center',out.width="75%")
```


