# Imagery Classification Testing

```{r}
library(stringr)
library(terra)
imgFolder <- paste0("S2A_MSIL2A_20210628T184921_N0300_R113_T10TGK_20210628T230915.",
                    "SAFE\\GRANULE\\L2A_T10TGK_A031427_20210628T185628")
img20mFolder <- paste0("~/sentinel2/",imgFolder,"\\IMG_DATA\\R20m")
imgDateTime <- str_sub(imgFolder,12,26)
imgDateTime
imgDate <- str_sub(imgDateTime,1,8)
```

```{r}
sentinelBands <- paste0("B",c("02","03","04","05","06","07","8A","11","12"))
sentFiles <- paste0(img20mFolder,"/T10TGK_",imgDateTime,"_",
                        sentinelBands,"_20m.jp2")
sentinel <- rast(sentFiles)
sentinelBands
names(sentinel) <- sentinelBands
```

**Spring**
```{r}
library(terra); library(stringr); library(igisci)
imgFolderSpring <- paste0("S2B_MSIL2A_20210603T184919_N0300_R113_T10TGK_20210603",
                     "T213609.SAFE\\GRANULE\\L2A_T10TGK_A022161_20210603T185928")
img10mFolderSpring <- paste0("~/sentinel2/",imgFolderSpring,"\\IMG_DATA\\R20m")
imgDateTimeSpring <- str_sub(imgFolderSpring,12,26)
```
**Summer**
```{r}
imgFolderSummer <- paste0("S2A_MSIL2A_20210628T184921_N0300_R113_T10TGK_20210628",
                      "T230915.SAFE\\GRANULE\\L2A_T10TGK_A031427_20210628T185628")
img10mFolderSummer <- paste0("~/sentinel2/",imgFolder,"\\IMG_DATA\\R20m")
imgDateTimeSummer <- str_sub(imgFolderSummer,12,26)
```

```{r}
sentinelBands <- paste0("B",c("02","03","04","05","06","07","8A","11","12"))
sentFilesSpring <- paste0(img10mFolderSpring,"/T10TGK_",
                              imgDateTimeSpring,"_",sentinelBands,"_20m.jp2")
sentinelSpring <- rast(sentFilesSpring)
names(sentinelSpring) <- paste0("spring",sentinelBands)

sentFilesSummer <- paste0(img10mFolderSummer,"/T10TGK_",
                              imgDateTimeSummer,"_",sentinelBands,"_20m.jp2")
sentinelSummer <- rast(sentFilesSummer)
names(sentinelSummer) <- paste0("summer",sentinelBands)
```

```{r}
RCVext <- ext(715680,725040,4419120,4429980)
sentRCVspring <- crop(sentinelSpring,RCVext)
sentRCVsummer <- crop(sentinelSummer,RCVext)
```


### Create a 18-band stack from both images
Since we have two imagery dates , we have a total of 18 variables to work with. So we'll put these together as a multi-band stack simply using `c()` with the various bands. 
```{r}
sentRCV <- c(sentRCVspring, sentRCVsummer)
```

```{r}
plotRGB(subset(sentRCV,c(7,3,2)),stretch="lin")
```

```{r}
plotRGB(subset(sentRCV,c(16,12,11)),stretch="lin")
```

### Extract the training data (20 m spring + summer)
We'll repeat some of the steps where we read in the training data and set up the land cover classes, reducing the 7 to 6 classes, then extract from our new 20-band image stack the training data. 
```{r}
library(igisci)
trainPolys <- vect(ex("RCVimagery/train_polys7.shp"))
ptsRCV <- spatSample(trainPolys, 1000, method="random")
LCclass <- c("forest","hydric","mesic","rocky","water","willow","xeric")
classdf <- data.frame(value=1:length(LCclass),names=LCclass)
trainPolys <- vect(ex("RCVimagery/train_polys7.shp"))
trainPolys[trainPolys$class=="willow"]$class <- "hydric"
ptsRCV <- spatSample(trainPolys, 1000, method="random")
LCclass <- c("forest","hydric","mesic","rocky","water","xeric")
classdf <- data.frame(value=1:length(LCclass),names=LCclass)

extRCV <- terra::extract(sentRCV, ptsRCV)[,-1] # REMOVE ID
sampdataRCV <- data.frame(class = ptsRCV$class, extRCV)
```

### CART model and prediction (20 m spring + summer)
\index{imagery classification from phenological change CART model 20 m spring + summer}
```{r}
library(rpart)
# Train the model
cartmodel <- rpart(as.factor(class)~., 
                   data = sampdataRCV, method = 'class', minsplit = 5)
```

The CART model tree for the two-season 20 m example (Figure \@ref(fig:imgCARTtreePheno20)), in contrast with the 10 m classification, does not end up with a few pixels classified as "rocky" (Figures \@ref(fig:imgCARTprobmapPheno20) and \@ref(fig:imgCARThighestPheno20)).  

```{r imgCARTtreePheno20, fig.cap="CART decision tree, Sentinel 20-m, spring and summer 2021 images"}
library(rpart.plot)
rpart.plot(cartmodel, fallen.leaves=F)
```

```{r imgCARTprobmapPheno20, fig.cap="CART classification, probabilities of each class, Sentinel-2 20 m, 2021 spring and summer phenology"}
CARTpred <- predict(sentRCV, cartmodel, na.rm = TRUE)
plot(CARTpred)
```

```{r}
LCpheno <- which.max(CARTpred)
cls <- names(CARTpred)
df <- data.frame(id = 1:length(cls), class=cls)
levels(LCpheno) <- df
```

```{r eval=F}
LCcolors <- c("green4","cyan","gold","grey90","mediumblue","red")
plot(LCpheno, col=LCcolors)
```

```{r imgCARThighestPheno20, fig.height=6.5, fig.width=6, fig.cap="CART classification, highest probability class, Sentinel-2 20 m, 2021 spring and summer phenology"}
library(tmap)
LCcolors <- c("green4","cyan","gold","grey90","mediumblue","red")
tm_shape(LCpheno) + 
    tm_raster(col="class",  
            col.scale=tm_scale_categorical(values=LCcolors),
            col.legend=tm_legend(frame=F)) +
  tm_graticules(alpha=0.5, col="gray")

```



We might compare this plot with the previous 20 m (Figure \@ref(fig:imgCARThighestProb20m)), 10 m (Figure \@ref(fig:imgCART10mHighestProb)), and 10 m spring+summer classification (\@ref(fig:imgCARThighestPheno)).


#### Cross validation : CART model for 20 m spring + summer
```{r}
set.seed(42)
k <- 5 # number of folds
j <- sample(rep(1:k, each = round(nrow(sampdataRCV))/k))
# table(j)
```

```{r}
x <- list()
for (k in 1:5) {
    train <- sampdataRCV[j!= k, ]
    test <- sampdataRCV[j == k, ]
    cart <- rpart(as.factor(class)~., data=train, method = 'class',
                  minsplit = 5)
    pclass <- predict(cart, test, na.rm = TRUE)
    # assign class to maximum probablity
    pclass <- apply(pclass, 1, which.max)
    # create a data.frame using the reference and prediction
    x[[k]] <- cbind(test$class, as.integer(pclass))
}
```

#### Accuracy metrics : CART model for 20 m spring + summer
Finally, as we did above, we'll look at the various accuracy metrics for the classification using the two-image (peak-growth spring and senescent summer) source.

**Confusion Matrix**
```{r}
y <- do.call(rbind, x)
y <- data.frame(y)
colnames(y) <- c('observed', 'predicted')
conmat <- table(y) # confusion matrix
```
```{r}
# change the name of the columns and rows to LCclass
if(length(LCclass)==dim(conmat)[2]){
  colnames(conmat) <- LCclass}else{
    print("Confusion matrix column dimension/LCclass array mismatch")
    print(str(LCclass))
    print(str(conmat))
  }
if(length(LCclass)==dim(conmat)[1]){
  rownames(conmat) <- LCclass}else{
    print("Confusion matrix row dimension/LCclass array mismatch")
    print(str(LCclass))
    print(str(conmat))
  }
```
```{r}
print(conmat)
```

**Overall Accuracy**
```{r}
n <- sum(conmat) # number of total cases
diag <- diag(conmat) # number of correctly classified cases per class
OA20ph <- sum(diag) / n # overall accuracy
OA20ph
```

**kappa**

```{r}
rowsums <- apply(conmat, 1, sum) 
p <- rowsums / n # observed (true) cases per class
colsums <- apply(conmat, 2, sum)
q <- colsums / n # predicted cases per class
expAccuracy <- sum(p*q)
kappa20ph <- (OA20ph - expAccuracy) / (1 - expAccuracy)
kappa20ph
```

**User and Producer accuracy**
```{r}
PA <- diag / colsums # Producer accuracy
UA <- diag / rowsums # User accuracy
if(length(PA)==length(UA)){
  outAcc <- data.frame(producerAccuracy = PA, userAccuracy = UA)
  print(outAcc)}else{
     print("length(PA)!=length(UA)")
     print(PA)
     print(UA)
  }

```
