# PointBlue Penguin Study

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

*Currently, this is the same as in the book, but will be expanded.*

### Interactive mapping of individual penguins abstracted from a big dataset

\index{big data abstraction}In a study of spatial foraging patterns by Adélie penguins (*Pygoscelis adeliae*), @Ballard2019 collected data over five austral summer seasons (parts of December and January in 2005-06, 2006-07, 2007-08, 2008-09, 2012-13) period using SPLASH tags that combine Argos satellite tracking with a time-depth recorder, with 11,192 observations. The 24 SPLASH tags were re-used over the period of the project with a total of 162 individual breeding penguins, each active over a single season. 

Our code takes advantage of the abstraction capabilities of base R and `dplyr` to select an individual penguin from the large data set and prepare its variables for useful display. Then the interactive mode of tmap works well for visualizing the penguin data -- both the cloud of all observations and the focused individual -- since we can zoom in to see the locations in context, and basemaps can be chosen. The `tmap` code colors the individual penguin's locations based on a decimal day derived from day and time. While interactive view is more limited in options, it does at least support a legend and title (Figure \@ref(fig:mapsPenguins)).

```{r echo=T}
set.seed(28)
```
```{r warning=F, message=F}
library(sf); library(tmap); library(tidyverse); library(igisci)
sat_table <- read_csv(ex("penguins/sat_table_splash.csv"))
obs <- st_as_sf(filter(sat_table,include==1), coords=c("lon","lat"), crs=4326)
uniq_ids <- unique(obs$uniq_id) # uniq_id identifies an individual penguin
onebird <- obs %>%
  filter(uniq_id==uniq_ids[sample.int(length(uniq_ids), size=1)]) %>%
  mutate(decimalDay = as.numeric(day) + as.numeric(hour)/24)
```
```{r eval=T}
tmap_mode = "view"
tmap_options(basemaps=c(Terrain = "Esri.WorldTerrain",
                        Imagery = "Esri.WorldImagery", 
                        NatGeo = "Esri.NatGeoWorldMap", 
                        Topo="OpenTopoMap",
                        Ortho="GeoportailFrance.orthos"))
penguinMap <- tm_basemap() +
  tm_shape(obs) + tm_symbols(col="black", border.lwd=0, size=0.01, alpha=0.33) + 
  tm_shape(onebird) + tm_symbols(col="decimalDay",
  palette="viridis", style="cont",n=8,size=0.6) +
  tm_legend() + 
  tm_layout() 
tmap_leaflet(penguinMap)
```

