# PointBlue Penguin Study

```{r include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

*Currently, this is the same as in the book, but will be expanded.*

### Interactive mapping of individual penguins abstracted from a big dataset

\index{big data abstraction}In a study of spatial foraging patterns by Adélie penguins (*Pygoscelis adeliae*), @Ballard2019 collected data over five austral summer seasons (parts of December and January in 2005-06, 2006-07, 2007-08, 2008-09, 2012-13) period using SPLASH tags that combine Argos satellite tracking with a time-depth recorder, with 11,192 observations. The 24 SPLASH tags were re-used over the period of the project with a total of 162 individual breeding penguins, each active over a single season. 

Our code takes advantage of the abstraction capabilities of base R and `dplyr` to select an individual penguin from the large data set and prepare its variables for useful display. Then the interactive mode of tmap works well for visualizing the penguin data -- both the cloud of all observations and the focused individual -- since we can zoom in to see the locations in context, and basemaps can be chosen. The `tmap` code colors the individual penguin's locations based on a decimal day derived from day and time. While interactive view is more limited in options, it does at least support a legend and title (Figure \@ref(fig:mapsPenguins)).

```{r echo=T}
set.seed(28)
```

```{r}
library(sf); library(tmap); library(tidyverse); library(igisci);
sat_table <- read_csv(ex("penguins/sat_table_splash.csv"))
obs <- st_as_sf(filter(sat_table,include==1), coords=c("lon","lat"), crs=4326)
uniq_ids <- unique(obs$uniq_id) # uniq_id identifies an individual penguin
onebird <- obs %>%
  filter(uniq_id==uniq_ids[sample.int(length(uniq_ids), size=1)]) %>%
  mutate(decimalDay = as.numeric(day) + as.numeric(hour)/24)

```


```{r mapsPenguins, echo=F, fig.cap=paste0("Observations of Adélie penguin migration from a 5-season study of a large colony at Ross Island in the SW Ross Sea, Antarctica; and an individual -- ",unique(onebird$uniq_id)," -- from season ",unique(onebird$season),". Data source: Ballard et al. (2019). Fine-scale oceanographic features characterizing successful Adélie penguin foraging in the SW Ross Sea. Marine Ecology Progress Series 608:263-277.")}
tmap_mode = "view"
penguinMap <- tm_basemap(server=c(Terrain = "Esri.WorldTerrain",
                Imagery = "Esri.WorldImagery",
                NatGeo  = "Esri.NatGeoWorldMap",
                Topo    = "OpenTopoMap",
                Ortho   = "GeoportailFrance.orthos")) +
  tm_shape(obs) +
    tm_dots(size=0.05, fill="black", fill_alpha=0.33) + 
  tm_shape(onebird) + 
    tm_dots(size=0.20,
       fill="decimalDay",
       fill.scale=tm_scale_continuous(values="tol.rainbow_pu_br"))
# tmap_leaflet(penguinMap, show=TRUE) # doesn't knit
penguinMap # works interactively, but knits to static map
```



