# PointBlue Penguin Study

```{r include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

*Currently, this is the same as in the book, but will be expanded.*

### Interactive mapping of individual penguins abstracted from a big dataset

#### Seabird Poisson Model

This \index{seabird poisson model}model is further developed in a separate case study where we'll map a prediction, but we'll look at the basics for applying a poisson family model in glm: *seabird observations*.  The data is potentially suitable since its counts with known time intervals of observation effort and initial review of the data suggests that the mean of counts is reasonably similar to the variance. The data were collected through the ACCESS program (@ACCESS) and a thorough analysis is provided by @Studwell2017.

```{r message=F, results='hide'}
library(igisci); library(sf); library(tidyverse); library(tmap)
library(maptiles)
transects <- st_read(ex("SFmarine/transects.shp"))
```

One of the seabirds studies, the black-footed albatross\index{black-footed albatross}, has the lowest counts (so a relatively rare bird\index{bird}), and studies of its distribution may be informative. We'll look at counts collected during transect cruises\index{bird transects} in July 2006 (Figure \@ref(fig:modAlbatrossCounts)).

```{r modAlbatrossCounts, message=F, warning=F, fig.cap="Black-footed albatross counts, July 2006"}
tmap_mode("plot")
transJul2006 <- transects %>%
  filter(month==7 & year==2006 & avg_tem>0 & avg_sal>0 & avg_fluo>0)
tm_basemap("Esri.NatGeoWorldMap") +
  tm_shape(transJul2006) + tm_symbols(fill="bfal", size="bfal",
                                      fill.scale=tm_scale_continuous(values="Greens"))
```

Prior studies have suggested that temperature\index{temperature}, salinity\index{salinity}, fluorescence\index{fluorescence}, depth\index{depth}, and various distances\index{distance (as explanatory variable)} might be good explanatory variables to use to look at spatial patterns, so we'll use these in the model. 

```{r}
summary(glm(bfal~avg_tem + avg_sal + avg_fluo + avg_dep +
            dist_land + dist_isla + dist_200m + dist_cord, 
            data=transJul2006, family=poisson))
```

We can see in the model coefficients\index{model coefficients} table several predictive variables that appear to be significant, which we'll want to explore further, and also map the results as we did with Sierra temperature data.

- avg_tem
- avg_sal
- avg_dep
- dist_land
- dist_200m


**Note** R will allow you to run glm on any family, but it's up to you to make sure it's appropriate. For example, if you provide a response variable that is *not* count data or doesn't fit the requirements for a poisson model, R will create a result but it's bogus. 

