
# 1. Initial data read and time series charting

Processing from Mason Kealy

```{r setup, include=FALSE}
library(readxl)
library(tidyverse)
library(lubridate)
```

### Reading in the Excel File

Mason: "This code is just reading in data from an Excel file I got from the Bioclimatology class last semester. I used sub-setting to remove the first row of the excel table it just had the units for each of the variables, keeping this would just mess up my code later on.
  This is what the tibble looked like when I first read in the excel data!"
```{r}
"RCV_carbon_flux_data.xlsx"
Carbon_Flux <- read_excel("GPP_PAR_data/Mountain meadow Carbon flux data.xlsx")
Carbonflux<- Carbon_Flux[-1, ]

print(Carbonflux)
```
  
You can see in the tibble above, that all the columns were in `chr` data type, this prevents the computer from seeing them as numbers.
  I started preparing the data for analysis by `mutate_all(as.numeric)` then setting the date column as a date with `as.Date` selected the variables I wanted and rounded the 7 digits values to get them looking a little nicer.
  Then I used `pivot_longer` to send the CO2 Flux Competent values to a new column and the names to a new column. This makes it a lot easier when creating a legend later on in Figure 1 (see next code chunk)
  *Then Jerry removed the zero GPP_sum values.* 
```{r}

Mountflux <- Carbonflux %>% 
mutate_all(as.numeric) %>% 
mutate(DOY = as.Date(Date,format = "%m-%d")) %>% 
select(Date:PAR_ave) %>% 
mutate(NEE = round(NEE_sum, digits = 3), 
       GPP = round(GPP_sum, digits = 3),
       RE = round(RE_sum, digits = 3)) %>%
#select(-NEE_sum, -GPP_sum, -RE_sum) %>% 
  pivot_longer(cols = c("NEE", "RE", "GPP"), names_to = "CO2_FLUX_COMPONENTS", values_to ="value") %>%
  filter(GPP_sum != 0, !is.na(PAR_ave))

print(Mountflux)


```

## Time series plot 
  Then I used ggplot to create a visualization of how NEE, GPP, an RE changed over a yearly time scale.
```{r}

Mountflux %>% 
 ggplot(aes(x = Date, y= value, color = CO2_FLUX_COMPONENTS))+
  geom_line()+
  geom_hline(yintercept = 0, linetype = "longdash", color = "black")+
  ggtitle(" Red Clover Valley: Interannual Components of Carbon Flux Measurements") +
  xlab("Day of Year") +
  ylab("g C/m^2")+
  labs(caption =
"Figure 1: Components of Carbon Flux measurements collected in 30 minute intervals in 2021
and added into ‘sum daily value’ , from flux tower in Red Clover Valley,California. Components included NEE (Net Ecosystem Exchange), RE (Respiration Efficiency), GPP (Gross Primary Productivity) all measured in g C m^2 and are plotted against an annual time scale with days as the unit of measure.")
  
```
 

# Non-linear model of PAR_ave ~ GPP_sum, using Michaelis-Menten model

*Not the right goal for a model between these variables, since GPP should derive from PAR, not the other way around.  But it does illustrate asymptotic data.*


First the scatter plot:

```{r}
plot(Mountflux$GPP_sum, Mountflux$PAR_ave,xlab="GPP_sum", ylab="PAR_ave")
```


### Try fitting a non-linear model (SSmicmen from base R)

Based on https://www.datacamp.com/tutorial/introduction-to-non-linear-model-and-insights-using-r

Using start values for Vm and K based on interpreting the graph. We'll use nonlinear least squares (`nls`) and starting values of the two coefficients.

```{r}
micmenMod <- nls(PAR_ave~(Vm)*GPP_sum/(K+GPP_sum), data=Mountflux, start=list(Vm=750,K=1))
summary(micmenMod)
```
**Calculate residual error**

```{r}
sse <- micmenMod$m$deviance()
null <- lm(PAR_ave~1, data=Mountflux)
sst <- data.frame(summary.aov(null)[[1]])$Sum.Sq
percent_variation_explained <- 100*(sst-sse)/sst
percent_variation_explained
```
```{r}
plot(Mountflux$GPP_sum, Mountflux$PAR_ave,xlab="GPP_sum", ylab="PAR_ave")
 lines(Mountflux$GPP_sum, 
       predict(micmenMod,Mountflux),col=2)
```

## polynomial models

```{r}
plot(Mountflux$PAR_ave,Mountflux$GPP_sum, xlab="PAR_ave", ylab="GPP_sum")

```

### poly 1

```{r}
ord = 1
summary(lm(GPP_sum ~ poly(PAR_ave, ord, raw=T), data=Mountflux))
ggplot(Mountflux, aes(PAR_ave, GPP_sum)) + geom_point() +
  stat_smooth(method=lm, formula=y~poly(x,ord,raw=T))

```



### poly 2

```{r}
ord = 2
summary(lm(GPP_sum ~ poly(PAR_ave, ord, raw=T), data=Mountflux))
ggplot(Mountflux, aes(PAR_ave, GPP_sum)) + geom_point() +
  stat_smooth(method=lm, formula=y~poly(x,ord,raw=T))

```


### poly 3


```{r}
ord = 3
summary(lm(GPP_sum ~ poly(PAR_ave, ord, raw=T), data=Mountflux))
ggplot(Mountflux, aes(PAR_ave, GPP_sum)) + geom_point() +
  stat_smooth(method=lm, formula=y~poly(x,ord,raw=T))
```

### poly 4
```{r}
ord = 4
summary(lm(GPP_sum ~ poly(PAR_ave, ord, raw=T), data=Mountflux))
ggplot(Mountflux, aes(PAR_ave, GPP_sum)) + geom_point() +
  stat_smooth(method=lm, formula=y~poly(x,ord,raw=T))
```

### poly 5
```{r}
ord = 5
summary(lm(GPP_sum ~ poly(PAR_ave, ord, raw=T), data=Mountflux))
ggplot(Mountflux, aes(PAR_ave, GPP_sum)) + geom_point() +
  stat_smooth(method=lm, formula=y~poly(x,ord,raw=T))
```

## power regression

```{r}
summary(lm(log(GPP_sum) ~ log(PAR_ave), data=Mountflux))
ggplot(Mountflux, aes(log(PAR_ave), log(GPP_sum))) + geom_point() +
  stat_smooth(method=lm)
```


