# CO2 flux data : Red Clover Valley

## 1. Initial data read and time series charting

Some ideas for processing are from Mason Kealy.  Data are from Andrew Oliphant.

```{r include=FALSE}
library(readxl)
library(tidyverse)
library(lubridate)
```

### Reading in the Excel File

First create a function to convert the first two rows into individual variables. We won't include the units.

```{r}
library(igisci); library(tidyverse); library(readxl)
```
```{r}
read_excel2 <- function(xl){
  vnames <- read_excel(xl, n_max=0) |> names()
  vunits <- read_excel(xl, n_max=0, skip=1) |> names()
  read_excel(xl, skip=2, col_names=vnames)
}
```


```{r}
Carbon_Flux <- read_excel2(ex("meadows/RCV_carbon_flux_data.xlsx"))
```
  
You can see in the tibble above, that all the columns were in `chr` data type, this prevents the computer from seeing them as numbers.
  I started preparing the data for analysis by `mutate_all(as.numeric)` then setting the date column as a date with `as.Date` selected the variables I wanted and rounded the 7 digits values to get them looking a little nicer.
  Then I used `pivot_longer` to send the CO2 Flux Competent values to a new column and the names to a new column. This makes it a lot easier when creating a legend later on in Figure 1 (see next code chunk)
  *Then Jerry removed the zero GPP_sum values.* 
```{r}

Mountflux <- Carbon_Flux %>% 
mutate_all(as.numeric) %>% 
mutate(DOY = as.Date(Date,format = "%m-%d")) %>% 
select(Date:PAR_ave) %>% 
mutate(NEE = round(NEE_sum, digits = 3), 
       GPP = round(GPP_sum, digits = 3),
       RE = round(RE_sum, digits = 3)) %>%
#select(-NEE_sum, -GPP_sum, -RE_sum) %>% 
  pivot_longer(cols = c("NEE", "RE", "GPP"), names_to = "CO2_FLUX_COMPONENTS", values_to ="value") %>%
  filter(GPP_sum != 0, !is.na(PAR_ave))

print(Mountflux)


```

## Time series plot 
  Then I used ggplot to create a visualization of how NEE, GPP, an RE changed over a yearly time scale.
```{r}

Mountflux %>% 
 ggplot(aes(x = Date, y= value, color = CO2_FLUX_COMPONENTS))+
  geom_line()+
  geom_hline(yintercept = 0, linetype = "longdash", color = "black")+
  ggtitle(" Red Clover Valley: Interannual Components of Carbon Flux Measurements") +
  xlab("Day of Year") +
  ylab("g C/m^2")+
  labs(caption =
"Figure 1: Components of Carbon Flux measurements collected in 30 minute intervals in 2021
and added into ‘sum daily value’ , from flux tower in Red Clover Valley,California. Components included NEE (Net Ecosystem Exchange), RE (Respiration Efficiency), GPP (Gross Primary Productivity) all measured in g C m^2 and are plotted against an annual time scale with days as the unit of measure.")
  
```
 

## Non-linear model of PAR_ave ~ GPP_sum, using Michaelis-Menten model

*Not the right goal for a model between these variables, since GPP should derive from PAR, not the other way around.  But it does illustrate asymptotic data.*


First the scatter plot:

```{r}
plot(Mountflux$GPP_sum, Mountflux$PAR_ave,xlab="GPP_sum", ylab="PAR_ave")
```


### Try fitting a non-linear model (SSmicmen from base R)

Based on https://www.datacamp.com/tutorial/introduction-to-non-linear-model-and-insights-using-r

Using start values for Vm and K based on interpreting the graph. We'll use nonlinear least squares (`nls`) and starting values of the two coefficients.

```{r}
micmenMod <- nls(PAR_ave~(Vm)*GPP_sum/(K+GPP_sum), data=Mountflux, start=list(Vm=750,K=1))
summary(micmenMod)
```
**Calculate residual error**

```{r}
sse <- micmenMod$m$deviance()
null <- lm(PAR_ave~1, data=Mountflux)
sst <- data.frame(summary.aov(null)[[1]])$Sum.Sq
percent_variation_explained <- 100*(sst-sse)/sst
percent_variation_explained
```
```{r}
plot(Mountflux$GPP_sum, Mountflux$PAR_ave,xlab="GPP_sum", ylab="PAR_ave")
 lines(Mountflux$GPP_sum, 
       predict(micmenMod,Mountflux),col=2)
```

## polynomial models

```{r}
plot(Mountflux$PAR_ave,Mountflux$GPP_sum, xlab="PAR_ave", ylab="GPP_sum")

```

### poly 1

```{r}
ord = 1
summary(lm(GPP_sum ~ poly(PAR_ave, ord, raw=T), data=Mountflux))
ggplot(Mountflux, aes(PAR_ave, GPP_sum)) + geom_point() +
  stat_smooth(method=lm, formula=y~poly(x,ord,raw=T))

```



### poly 2

```{r}
ord = 2
summary(lm(GPP_sum ~ poly(PAR_ave, ord, raw=T), data=Mountflux))
ggplot(Mountflux, aes(PAR_ave, GPP_sum)) + geom_point() +
  stat_smooth(method=lm, formula=y~poly(x,ord,raw=T))

```


### poly 3


```{r}
ord = 3
summary(lm(GPP_sum ~ poly(PAR_ave, ord, raw=T), data=Mountflux))
ggplot(Mountflux, aes(PAR_ave, GPP_sum)) + geom_point() +
  stat_smooth(method=lm, formula=y~poly(x,ord,raw=T))
```

### poly 4
```{r}
ord = 4
summary(lm(GPP_sum ~ poly(PAR_ave, ord, raw=T), data=Mountflux))
ggplot(Mountflux, aes(PAR_ave, GPP_sum)) + geom_point() +
  stat_smooth(method=lm, formula=y~poly(x,ord,raw=T))
```

### poly 5
```{r}
ord = 5
summary(lm(GPP_sum ~ poly(PAR_ave, ord, raw=T), data=Mountflux))
ggplot(Mountflux, aes(PAR_ave, GPP_sum)) + geom_point() +
  stat_smooth(method=lm, formula=y~poly(x,ord,raw=T))
```

## power regression

```{r}
powerModel <- lm(log(GPP_sum) ~ log(PAR_ave), data=Mountflux)
summary(powerModel)
ggplot(Mountflux, aes(log(PAR_ave), log(GPP_sum))) + geom_point() +
  stat_smooth(method=lm)
```
Now plot without the log transformed axes, we'll need to create the curve with the model, using the coefficients.

```{r}
coeff = exp(powerModel$coefficients[1])
power = powerModel$coefficients[2]
fit_func <- function(x) coeff*x^power

fluxPlot <- ggplot(Mountflux) +
  geom_point(aes(x=PAR_ave,y=GPP_sum))
fluxPlot <- fluxPlot +
  stat_function(fun=fit_func, color="orange",lwd=2, alpha=0.5)
fluxPlot
```

```{r}
fit_eq = paste("y = ", round(coeff,4), " x^", round(power,3))
fluxPlot <- fluxPlot +
  annotate("text",x=200,y=40,label=fit_eq,size=6)
fluxPlot
```

